---
title: "TLC Data Analysis"
author: "Alex Powers"
date: "October 2018"
output:
  html_document: default
  pdf_document: default
---

### Introduction

This is an examination of the 2013 data on NYC yellow cabs. This dataset, originally shared with the public by [Chris Whong](https://chriswhong.com/open-data/foil_nyc_taxi/), doesn't include the boro (green) cabs that were introduced in the summer of 2013 and forbidden from picking up passengers in the "Manhattan Exclusionary Zone." Sounds dystopic... 

What follows is freewheeling exploration of interesting facets of the yellow cab data. 

### Setup
Initial setup of the TLC data was done on Google BigQuery, detailed [elsewhere](https://github.com/al3xpowers/tlc-data-analysis/blob/master/tlc1.sql). Connecting BigQuery to R required some configuration of the BigQuery API.
First we sample 850K records for analysis, or ~.05% of the total for this exploratory exercise.
We clean up the data by only taking payments done with card (since cash tips are't recorded),
as well as excluding trips of extreme length (>= 100 miles). Note that 100 miles is enough to reach nearly all of the Hamptons from Manhattan.

```{r message=FALSE, warning=FALSE}
setwd("~/Workdir/tlc/R")
library(bigrquery)
library(ggplot2)
library(tidyverse)
project = "sylvan-rampart-159916"
set_service_token("zuul.json")
sql <- "SELECT trip_distance, tip_amt, rand(5) AS rand FROM tlc.combo7
WHERE payment_type = 'CRD' AND trip_distance < 100
ORDER BY rand LIMIT 850000"
tdX <- query_exec(sql, project = project, useLegacySql = FALSE, max_pages = Inf)
```


### Does trip distance increase tipping? 
We'd expect to see that drivers tip more as rides get longer - is that the case? 

```{r}
tdXlm <- lm(tip_amt ~ trip_distance, data = tdX)
summary(tdXlm)
```

We can see that there is a positive and strong effect of trip distance on tip amount to the point where trip distance explains roughly half the variability in tip amount.

Plotting these data reveals additional characteristics. 

```{r}
ggplot(tdX, aes(trip_distance, tip_amt)) +
  geom_point(shape = 20, alpha = 1/16) +
  geom_smooth(method = lm) +
  coord_cartesian(xlim = c(0, 25), ylim = c(0, 25)) +
  theme_bw() +
  labs(x = "Trip distance (mi)", y = "Tip amount ($)")
```

The trendline, plotted, clearly follows the bulk of the points. You can see some striation in the scatter, which indicates that people often round tips to the nearest whole dollar amount. Oftentimes the tip is $0. There's another interesting angle we could explore here - for instance, what is responsible for the distinct "tail" of points that shoots out at a sharp upward trajectory near the origin and above the trendline? Let's save that for another day..

### Does trip distance increase tipping generosity?

It's fairly unsurprising that increased trip distance leads to larger trips. But do riders respond by generously rewarding drivers for long-distance hauls, or does sticker shock make them more likely to economize on the tips? 

First we yank down some data on total fare as well as tip paid.

```{r}
sql2 <- "SELECT trip_distance, tip_amt, total_amt, passenger_count, rand(5) AS rand FROM tlc.combo7
WHERE payment_type = 'CRD' AND trip_distance < 100
ORDER BY rand LIMIT 850000"
tdY <- query_exec(sql2, project = project, useLegacySql = FALSE, max_pages = Inf)
```

The next step is to recalculate the tip amount as a share of total fare (let's call this *tip yield*) variable and run a regression with that as the outcome.

```{r}
library(dplyr)
tdYA <- tdY %>% 
  mutate(tip_yield = tip_amt/total_amt) %>% select(-rand)

tdYAlm <- lm(tip_yield ~ trip_distance, data = tdYA)
summary(tdYAlm)
```

Now we see a slight, but statistically significant, *negative* relationship. People become slightly less generous as the meter climbs.

```{r warning = FALSE}
ggplot(tdYA, aes(trip_distance, tip_yield)) +
  geom_point(shape = 20, alpha = 1/16) +
  geom_smooth(method = lm) +
  coord_cartesian(xlim = c(0, 25), ylim = c(0, 1)) +
  theme_bw() +
  labs(x = "Trip distance (mi)", y = "Tip yield (%)")
```

It's a bit difficult to parse the trends in this scatter - it's not as clear as the last one. Let's take a different approach by splitting the dataset into four sections by ride length and eyeball any differences in the distributions.

```{r message=FALSE, warning=FALSE}
tdY1 <- tdYA %>% filter(trip_distance > 0 & trip_distance <= 5) %>% 
  select(trip_distance, passenger_count, tip_yield)
tdY2 <- tdYA %>% filter(trip_distance > 5 & trip_distance <= 15) %>% 
  select(trip_distance, passenger_count, tip_yield)
tdY3 <- tdYA %>% filter(trip_distance > 15 & trip_distance <= 25) %>% 
  select(trip_distance, passenger_count, tip_yield)
tdY4 <- tdYA %>% filter(trip_distance > 25) %>% 
  select(trip_distance, passenger_count, tip_yield)

q1 <-ggplot(tdY1, aes(tip_yield)) +
  geom_histogram(color="black", alpha = .4, fill="#FF6666", binwidth = .025) +
  coord_cartesian(xlim = c(0, 1)) +
  theme_bw() +
  coord_cartesian(xlim = c(0, 1)) +
  labs(title = "0-5 mi, N = 713,718")
q2 <-ggplot(tdY2, aes(tip_yield)) +
  geom_histogram(color="black", alpha = .4, fill="#FF6666", binwidth = .025) +
  coord_cartesian(xlim = c(0, 1)) +
  theme_bw() +
  labs(title = "5-15 mi, N = 111,015")
q3 <-ggplot(tdY3, aes(tip_yield)) +
  geom_histogram(color="black", alpha = .4, fill="#FF6666", binwidth = .025) +
  coord_cartesian(xlim = c(0, 1)) +
  theme_bw() +
  labs(title = "15-25 mi, N = 20,395")
q4 <-ggplot(tdY4, aes(tip_yield)) +
  geom_histogram(color="black", alpha = .4, fill="#FF6666", binwidth = .025) +
  coord_cartesian(xlim = c(0, 1)) +
  theme_bw() +
  labs(title = "25+ mi, N = 533")

library(gridExtra)
grid.arrange(q1, q2, q3, q4, nrow = 2)
```

The share of people dispensing little to no tip increaess dramatically in the bucket of rides over 25 miles. You also see, for this distance range, that nobody's tipped over 75% of the nonoptional cost of the trip - not so for the other 3 tranches. 

### What other factors influence tipping behavior?

Let's test against the other data in the TLC release by running a multiple linear regression on all plausibly linked variables, with the response variable as tip yield.
```{r}
sql3 <- "SELECT trip_distance, tip_amt, total_amt, pickup_datetime, passenger_count, trip_time_in_secs, mta_tax, surcharge, tolls_amt, vendor_id, hack_license, rand(5) AS rand FROM tlc.combo7
WHERE payment_type = 'CRD' AND trip_distance < 100
ORDER BY rand LIMIT 850000"

tdZ <- query_exec(sql3, project = project, useLegacySql = FALSE, max_pages = Inf)
tdZlm <- lm (tip_amt/total_amt ~ trip_distance + pickup_datetime + passenger_count +
               trip_time_in_secs + mta_tax + surcharge + tolls_amt + vendor_id, data = tdZ)
```
```{r}
summary (tdZlm)
```

And let's get standardized coefficients so we can compare variable effects easily. 

```{r, message=FALSE, warning=FALSE}
library(QuantPsyc)
lm.beta(tdZlm)
```

A good many factors significantly contribute to determining tip yield: trip distance, tolls, MTA tax, pickup time, and surcharges. The aforementioned variables were ordered by decreasing strength of their effect, with trip distance head and shoulders above the rest. The vendor of the payment system also has a statistically significant effect – but being a categorical variable, its effect strength can’t be assessed in tandem with the others. It is the case, however, that trips using the vendor system CMT have, on average, tip yields that are 0.6 percentage points higher than those drivers using VTS. This is a significant financial difference between the two systems, and when I researched this oddity I found that this has also been [pointed out](http://iquantny.tumblr.com/post/107245431809/how-software-in-half-of-nyc-cabs-generates-52) by Ben Wellington of IQuantNY fame. The CMT system's calculation of recommended tips included taxes and tolls, juicing their numbers.

#### Does weather influence tipping?

##### Does inclement weather have an effect? 

Anecdotally speaking, taxis become scarce (and ride-share prices surge) during rain squalls - but can nasty weather's effect on tipping behavior be detected in the data we have at hand?

Using Weather Underground’s entire 2013 records from the Central Park station, I found that the 9 days with both rain and snow saw a statistically significant increase in tip yield for those days – albeit a small one, of 0.2 percentage points. Neither rain nor snow, in isolation, has a significant effect. The boxplot below illustrates this neatly. 

```{r}
weather <- read.csv("unified-weather-labels.csv", header = TRUE) 
w1lm <- lm (tip_share ~ events, data = weather)
summary(w1lm)
ggplot(weather, aes(events, tip_share)) +
  geom_boxplot(aes(fill=events)) +
  theme_bw() +
  labs(x = "Weather conditions", y = "Tip yield (% of total fare)") +
  guides(fill = FALSE)
```

##### Does temperature have an effect? 

It's straightforward to analyze this - just use a simple linear regression.

```{r}
w2lm <- lm (tip_share ~ avg_temp, data = weather)
summary(w2lm)

ggplot(weather, aes(avg_temp, tip_share)) + 
  geom_point(shape = 20, alpha = 1/4) +
  geom_smooth(method = lm, se=FALSE) +
  theme_bw() +
  labs(x = "Daily Average Temperature (°F)", y = "Tip yield (% of total fare)")
```

Temperature has a statistically significant effect on tipping; a one-degree increase in temperature results in a 0.004 percentage point decrease in tip yield. This tendency is fairly apparent in the scatter plot data. Perhaps people are more appreciative of warm cabs in winter conditions?

### Are some drivers better than others at generating generous tips?

We saw that tipping generosity generally decreases as trip distance lengthens. But what if there were a different class of drivers that bucked this trend? 

A simple answer could be achieved by calculating the Pearson correlation coefficient of the tip amount <> trip distance relationship for every driver, and look at how those correlations vary, but this misses detail that can help answer the question.

Instead, let's model driver characteristics by performing a linear regression for the tip amount <> trip distance relationship for each driver - and follow it up with cluster analysis on these driver-level regressions to identify groupings of drivers. Note that I'm going to focus not on tip as a yield of the overall fare, but the raw tip amount. 

```{r, message=FALSE, warning=FALSE}
library(factoextra)
library(nlme)
tdZA <- tdZ %>% 
  mutate(drivid = group_indices_(., .dots = "hack_license")) # adding this index improves query performance
tdZA <- tdZA %>% group_by(drivid) %>% filter(n()>19) # removing drivers with less than 20 rides
modelz <- lmList(tip_amt ~ trip_distance | drivid, data = tdZA)
```

Now we extract the characteristics that we're interested in from this crop of 22,129 driver regressions. I'm going to snag the intercept coefficient, the trip distance variable coefficient, the R-squared, the p-value for the intercept coefficient, and the p-value for the trip distance coefficient.

```{r}
output1 <- function(x) c(coef(x), summary(x)$r.squared, 
                         summary(x)$coefficients[,4])
output2 <- t(sapply(modelz,output1))
colnames(output2) <- c("intercept", "trip_distance", "r_squared", "intercept_p", "trip_distance_p")
output2 <- as.data.frame(output2)
sig_output <- output2 %>% filter(trip_distance_p<0.05) # filtering out drivers that don't have a significant trip distance coefficient
```

I'm choosing to use k-means clustering to identify groupings based on two important and interpretable variables: the R-squared values and the trip distance coefficient (slope) values for each driver regression. In order to select the value of *k*, we will use the gap statistic in the manner recommended by [Tibshirani et al. (2001)](https://statweb.stanford.edu/~gwalther/gap).


```{r message=FALSE, warning=FALSE}
library(cluster)
sig_scaled <- scale(sig_output)
set.seed (72)
CG <- clusGap(sig_scaled[,2:3], FUNcluster = kmeans, nstart = 25, d.power = 2, K.max = 10, B = 200)
k_choice <- maxSE(CG$Tab[,"gap"], CG$Tab[,"SE.sim"], method="Tibs2001SEmax")
k_choice
```

The outcome is that we should use a *k* of 3. Although the clustering itself is performed on the normalized data, the cluster assigned to each driver is painted on to the original dataset in the service of interpetability.

```{r warning=FALSE}
library(data.table)
sig_scaled <- data.frame (sig_scaled, stringsAsFactors = F)
centers <- kmeans(sig_scaled[,2:3], centers = 3, nstart = 25)$centers
centers.dt <- data.table(centers, key="trip_distance") # this is necessary to ensure the clusters are consistently ordered (i.e. after each knit cluster 1 remains cluster 1, not 3, etc.)
km2 = kmeans(sig_scaled[,2:3], centers = centers.dt, iter.max = 30) 
km2$cluster <- as.factor(km2$cluster)
ggplot(sig_output, aes(trip_distance, r_squared, color = km2$cluster)) + 
  geom_point() +
  theme_bw() +
  labs(title = "Clustering of Drivers' Trip Distance <> Tip Amt Relationship", x = "Trip Distance Coefficient",
       y = "R-Squared", color = "Cluster")
sig_output <- cbind(sig_output, km2$cluster)
sig_output %>% group_by(km2$cluster) %>% summarize(n = n(), trip_dist_mean = mean(trip_distance), r_sq_mean = mean(r_squared))
```

The 61 drivers in cluster 3 are far superior to those in clusters 1 and 2 in terms of consistently converting trip distance into unusually large tips, with a mean trip distance coefficient of 5.5 and mean R-squared of 0.71. The 20,374 drivers in clusters 1 and 2 do not consistently convert trip distance into whopping tips.

This could serve as a jumping-off point for deeper dives into what distinguishes the drivers of cluster 1 – drivers whose distance from the rest of the pack would have been subsumed in a simpler Pearson-only analysis. 

Perhaps these exceptional drivers are better at securing rides to and from destinations with attributes that make them unusually profitable. Hamptons – NYC trips might qualify, as might trips from 5-star hotels. These drivers could also be providing a high level of service – maybe they’re really good at reading customers’ preferences for engaging conversation, comfortable silence, or a driving style that's only moderately aggressive.

### Wrap-Up

Here are some headline takeaways from what I've seen so far:

1. Greater trip distance leads to higher tips... but decreased tip generosity.

2. The payment systems installed in cabs use different methods for calculating tips, which leads to significantly different tip yields.

3. Riders tip more generously on days with both rain and snow.. and when temperatures drop.

4. A subset of drivers (roughly 0.3%) appear to be consistently successful at bagging large tips.


```{r warning=FALSE, include=FALSE}
library(xfun)
optipng(dir = "/Users/AUGUSTUS/Workdir/tlc/R/apowMD1_files") 
#installed via homebrew; compresses image size
```

